---
title: "Finite mixture regression"
author:
  - Yuance He
date: "10/12/2018"
documentclass: article
papersize: letter
fontsize: 11pt

output: pdf_document
---

# 1. E-step

\begin{align*}
Q(\Psi | \Psi^{(k)}) &= E[l_n^c(\Psi)|(x,y)]\\
&= \sum_zP(z|x, y, \Psi^{(k)}) l_n^c(\Psi)\\
&= \sum_zP(z|x, y,\Psi^{(k)})\sum_{i=1}^n\sum_{j=1}^m z_{ij}\{ln\pi_j+ln\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}\\
&= \sum_{i=1}^n\sum_{j=1}^m [\sum_z z_{ij}P(z|(x, y), \Psi^{(k)})] 
\{ln\pi_j+ln\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}\\
&= \sum_{i=1}^n\sum_{j=1}^m \{ln\pi_j+ln\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}
E(z_{ij}|(x, y), \Psi^{(k)})\\
\end{align*}
Since we have $p_{ij}^{(k+1)}=\sum_{z_{ij}}p(z_{ij}|x_i,y_i,\Psi^{(k)})z_{ij}=E[z_{ij}|x_i,y_i,\Psi^{(k)}]=\frac{\pi_j^{(k)}\phi(y_i-x_i^T\beta_j^{(k)};0,\sigma^{2^{(k)}})}{\sum_{j=1}^m\pi_j^{(k)}\phi(y_i-x_i^T\beta_j^{(k)};0,\sigma^{2^{(k)}})}$,
\begin{align*}
Q(\Psi | \Psi^{(k)}) =\sum_{i=1}^n\sum_{j=1}^m p_{ij}^{(k+1)}\{ln\pi_j+ln\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}\\
\end{align*}

# 2. M-step

Now we need to maxmize $Q(\Psi | \Psi^{(k)})$.
\begin{align*}
Q(\Psi | \Psi^{(k)}) &=\sum_{i=1}^n\sum_{j=1}^m p_{ij}^{(k+1)}\{ln\pi_j+ln\varphi(y_i-x_i^T\beta_j;0, \sigma^2)\}\\
Q(\Psi | \Psi^{(k)}) &=\sum_{i = 1}^{n} \sum_{j = 1}^{m}p_{ij}^{(k+1)}ln\pi_j^{(k)}+ \sum_{i = 1}^{n} \sum_{j = 1}^{m} p_{ij}^{(k+1)} ln(\frac{1}{\sigma\sqrt{2 \pi} })+\sum_{i = 1}^{n} \sum_{j = 1}^{m} p_{ij}^{(k+1)} (- \frac{(y_i - x_i^T \beta_j^k)^2}{2\sigma^2})\\
\end{align*}

## (1) $\pi_j^{k+1}$
\begin{align*}
\frac{\partial Q}{\partial \pi_j} &=\frac{\partial}{\partial \pi_j} \sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}\ln\pi_j\\
&=\ln\pi_1\sum_{i=1}^{n}p_{i1}^{(k+1)}+...+\ln\pi_j\sum_{i=1}^{n}p_{ij}^{(k+1)}+...\ln\pi_m\sum_{i=1}^{n}p_{im}^{(k+1)}\\
&=\frac{\sum_{i=1}^{n}p_{ij}}{\pi_j}-\frac{\sum_{i=1}^{n}p_{im}}{\pi_m}\\
&=0\\
\end{align*}
Let $\frac{\sum_{i=1}^n p_{ij}^{k+1}}{\pi_j}=n$, then $\pi_j^{k+1} =\frac{\sum_{i=1}^n p_{ij}^{k+1}}{n}$

## (2) $\beta _j^{k+1}$
\begin{align*}
\frac{\partial Q}{\partial\beta_j} &=\frac{\partial}{\partial \beta_j} \sum_{i=1}^n \sum_{j=1}^m p_{ij}^{k+1}\frac{(y_i-x_i^T \beta_j)^2}{2\sigma^2} \\
&=-\frac{1}{2\sigma^2}\sum_{i=1}^{n}p_{ij}^{(k+1)}(-2)(y_i-x_i^T\beta_j^{(k+1)})\frac{\partial (x_i^T\beta_j^{(k+1)})}{\partial \beta_j}\\
&=0\\
\beta_j^{(k+1)}&=(\sum_{i=1}^{n}x_ix_i^Tp_{ij}^{(k+1)})^{-1}(\sum_{i=1}^{n}x_ip_{ij}^{(k+1)}y_i)
\end{align*}

## (3) $\sigma^{2(k+1)}$
\begin{align*}
\frac{\partial Q}{\partial \sigma^{2(k+1)}}&=\frac{\partial}{\partial\sigma^{2(k+1)}} (\frac{1}{2} \sum_{i=1}^n \sum_{j=1}^m p_{ij}^{k+1}\log(2 \sigma^2 \pi)) -( \sum_{i=1}^n \sum_{j=1}^m p_{ij}^{k+1}\frac{(y_i-x_i^T \beta_j)^2}{2\sigma^2})\\
&=0\\
n &=\frac{\sum_{i=1}^{n}\sum_{j=1}^{m}p_{ij}^{(k+1)}(y_i-x_i^T\beta_j^{(k+1)})^2}{\sigma^{2(k+1)}}\\
\sigma^{2(k+1)}&=\frac{ \sum_{i=1}^n \sum_{j=1}^m p_{ij}^{k+1}(y_i-x_i^T \beta_j^k)^2}{n}\\
\end{align*}






# 3. Implementation of the Algorithm

```{r}
regmix_em <- function(y, xmat, pi.init, beta.init, sigma.init, 
                      control=list(maxit = 500, tol =1e-5)){
  r <- nrow(xmat)
  c <- ncol(xmat)
  l <- length(pi.init)
  max <- control$maxit
  xmat <- as.matrix(xmat)
  tol <- control$tol
  pi <- pi.init
  beta <- beta.init
  sigma <- sigma.init
  p <- matrix(0,r,l)
  beta.new=matrix(0,c,l)
  for (k in 1:max) {
    for (i in 1:n) {
      p[i,] <- pi * dnorm(y[i] - xmat[i,] %*% beta, mean=0, sigma)/
           sum(pi * dnorm(y[i] - xmat[i,] %*% beta, mean=0, sigma))
    }
    pi.new <- colMeans(p)
    
    for (j in 1:l) {
      beta.new[,j] <- solve(t(xmat)%*%diag(p[,j])%*%xmat)%*%t(xmat)%*%diag(p[,j])%*%y
    }
    
    sigma.new <- sqrt(sum(p * (y %*% t(rep(1, l)) - xmat %*% beta.new)^2)/n)
    
    dv <- sum(abs(pi.new-pi))+sum(abs(beta.new-beta))+abs(sigma.new-sigma)
    if(dv < tol) break
    
    pi <- pi.new
    beta <- beta.new
    sigma <- sigma.new
    
  }
  
  if(k == max)
  print("Maximum Iteration")
  
  list(pi = pi.new, beta = beta.new, sigma = sigma.new, dv = dv, iter = k)
}
```






# 4. Generating Data

```{r}
regmix_sim <- function(n, pi, beta, sigma) {
    K <- ncol(beta)
    p <- NROW(beta)
    xmat <- matrix(rnorm(n * p), n, p) # normal covaraites
    error <- matrix(rnorm(n * K, sd = sigma), n, K)
    ymat <- xmat %*% beta + error # n by K matrix
    ind <- t(rmultinom(n, size = 1, prob = pi))
    y <- rowSums(ymat * ind)
    data.frame(y, xmat)
}
n <- 400
pi <- c(.3, .4, .3)
bet <- matrix(c( 1,  1,  1, 
                -1, -1, -1), 2, 3)
sig <- 1
set.seed(1205)
dat <- regmix_sim(n, pi, bet, sig)
result<- regmix_em(y = dat[,1], xmat = dat[,-1], 
             pi.init = pi / pi / length(pi),
             beta.init = bet * 0,
             sigma.init = sig / sig, 
             control = list(maxit = 500, tol = 1e-5))
result
```





                     
                